Exploiting First-Class Arrays in Fortran for Accelerator Programming

### 2010-12-03 opencl/papers/foropencl.tex
PGI
more

@@

\begin{enumerate}
\item Elemental functions
\item Pure procedures
\item Shift operators
\item Array notation
\end{enumerate}

@@

\cite{chamberlain04zpl}

>>>>

%%
%%
%%

@@

\subsubsection*{Elemental functions}

An elemental function consumes and produces scalar values, but can be
applied to variables of array type such that the function is
applied to each and every element of the array.  This allows programmers
to avoid explicit looping and instead simply state that they intend a
function to be applied to every element of an array in parallel, deferring
the choice of implementation technique to the compiler.  Elemental
functions are intended to be used for data parallel programming, and
as a result must be side effect free (or, \emph{pure}).

\subsubsection*{Pure procedures}

Pure procedures, like elemental functions, must be free of side effects.
The absence of side effects removes ordering constraints from the compiler
allowing it to invoke pure functions out of order.  Procedures and functions
of this sort are also common in pure functional languages like Haskell,
and are exploited by compilers in order to emit parallel code automatically
due to their suitability for compiler-level analysis.

\subsubsection*{Shift operators}

Many array-based algorithms require the same operation to be performed
on each element of the array using the value of that element and some small
set of neighboring cells.  Often programmers implement these operations that
are local to each element within the inner-loop of a set of nested FOR-
or DO-loops using offsets relative to the current array index.  An alternative
to this local-view of the algorithm is to take a global view and write the
algorithm in terms of the whole array.  For example, consider a 1D array in
which we wish to subtract the $(i-1)$th element from the $i$th for all
elements.  One way to look at this is that we are subtracting the entire array
shifted by one element from itself.

Fortran provides a set of shifting operators that allow programmers to
define operations based on shifted arrays.  These intrinsic operators take
an array, a dimension, and the amount by which it should be shifted (using
the sign to indicate direction).  By defining operations on entire arrays
based on a global view of them shifted relative to each other, programmers can
avoid explicit looping and potentially tricky index arithmetic.  Furthermore,
analysis of the extent of the set of shifted arrays in a given expression
allows analysis tools to determine the amount of temporary or buffer storage
necessary to hold intermediate values during whole array operations.  With
explicit loops, programmers must maintain this temporary storage manually.

\subsubsection*{Array notation}

The final piece that we exploit is the array notation present in modern
versions of Fortran.  ....

@@

\cite{chamberlain04zpl,pgi10accelerator}
